Do early environmental experiences impact numerical discrimination abilities in a lizard?

### 2025-02-28 ms.qmd
Updates
updates

## Abstract
From social behaviour to navigating complex environments, quantitative abilities can be crucial to fitness. However, early life conditions can impact brain development to affect quantitative competence. For example, early thermal conditions or glucocorticoid concentrations (GCs) – stress-related hormones transmitted by the parents – can play a prominent role in shaping cognition through their effects on brain physiology. Furthermore, temperature can lead to increased levels of GCs that can be transmitted to offspring, emphasizing the need to study the combined effects of these two factors. Here, we investigated the effects of elevated prenatal corticosterone (CORT) – the main GC in reptiles – and incubation temperature on numerical discrimination in the common garden skink (_Lampropholis guichenoti_). We assessed numerical abilities through a spontaneous choice test with food as a stimulus. Employing a repeated measures design, we subjected lizards to five numerical tests each differing in the ratios between the two choices (1 VS 4, 1 VS 3, 2 VS 4, 2 VS 3, 3 VS 4). Contrary to our predictions, we found no evidence for treatment effects on lizard behaviour and no use of numerical discrimination during foraging in _L. guichenoti_. Our findings build on previous studies suggesting that reptiles feeding on live prey may have difficulty distinguishing between different prey quantities. Despite our results, further investigation of numerical abilities in live-prey-feeding reptiles and the impact of early conditions on other taxa is warranted, as these areas remain understudied.

## Significance statement
Discriminating between quantities can be crucial for individual fitness, yet the impact of early environmental factors on the development of quantitative competence remains largely unexplored. In this study, we investigated, for the first time, the effects of prenatal corticosterone exposure and incubation temperature on numerical discrimination in a reptile. Contrary to our predictions, we did not observe any evidence of numerical discrimination and therefore no impact of early developmental environment. Our results align with previous studies showing that reptiles consuming live prey struggle to differentiate prey numbers in spontaneous choice tests. These results open new avenues for exploring the evolution of numerical abilities in reptiles and identifying factors influencing the development of quantitative skills in this group. Further research should explore alternative methods, such as trained tasks, to better understand the cognitive mechanisms underlying quantitative competence in reptiles.

@@

Numerical competence is often evaluated by considering the types of cues animals use [@plotnik2019elephants; @cooper2024tell], the ecological contexts in which these abilities are deployed [@mccomb1994roaring; @chittka1995can; @uller2003salamanders; @carazo2009quantity], and the neural mechanisms that underlie them [@nieder2005counting]. The latter have demonstrated that non-verbal numerical representations in mammals activates an extensive neural pathway including regions involved in eye movement, attention control, or complex learning [@nieder2005counting]. The connection between numerical discrimination and brain physiology underscores the importance of understanding how brain development influences numerical abilities.

Brains are particularly sensitive to environmental inputs during early stages of development [@coomber_independent_1997; @zhu_prenatal_2004], with long-lasting effects on cognitive abilities [@zhu_prenatal_2004; @amiel_egg_2014]. For instance, glucocorticoids (GCs), steroid hormones that mediate the physiological stress response [@sapolsky_how_2000], can alter performance in spatial, associative, or reversal learning tasks if elevated during early stages of development [@szuran_water_1994; @zhu_prenatal_2004; @crino_corticosterone_2014-learn; @farrell_developmental_2015-learn; @bebus_associative_2016; @lui2017chronic]. In ectotherms, temperature can affect GC levels in parents, cascading into effects on the developing offspring [i.e., maternal effects; @jessop2016multiscale; @racic_effects_2020; @mentesana2022glucocorticoids; @Crino_2023]. Prenatal exposure to GCs is typically associated with impaired learning skills, but these effects have been shown to vary depending on factors such as dose, sex, or type of cognitive test employed [@szuran_water_1994; @farrell_developmental_2015-learn; @bebus_associative_2016; @haq_prenatal_2021]. Most past studies have focused on the effects of GC exposure during development on spatial, associative, or reversal learning. Few studies have focused on how exposure to GCs during development can affect quantitative abilities, despite the importance of numerical discrimination.

Although maternally transmitted GCs are a significant factor in cognitive development, other environmental conditions can also influence cognition, either independently or through interactions with GCs. For instance, the early thermal environment plays a prominent role in brain development and cognitive abilities in ectotherms [@coomber_independent_1997;@amiel_hotter_2012; @amiel_egg_2014; @clark_colour_2014; @amiel_effects_2017; @vila_pouca_incubation_2018; @vila_pouca_quantity_2019]. In lizards, incubating eggs at elevated temperatures improves cognitive abilities consistent with elevated neuronal density in some areas of the brain [@amiel_hotter_2012; @amiel_egg_2014; @clark_colour_2014; @amiel_effects_2017], but impairs cognition at extreme temperatures [@dayananda_incubation_2017; @abayarathna_effects_2020]. In addition, when @vila_pouca_quantity_2019 tested the effects of elevated incubation temperature on numerical discrimination in  Port-Jackson sharks (_Heterodontus portusjacksoni_) and found that sharks incubated at elevated temperatures learned faster in a quantity discrimination task [@vila_pouca_quantity_2019]. However, the effects of early thermal environment on numerical abilities in other taxa remains largely unexplored.

Early-life GCs and thermal conditions could have interactive effects that influence cognitive development. Understanding how these factors influence quantitative abilities could reveal important insights into the ecological relevance of numerosity. Given the influence of numerical cognition on fitness, we would expect evolutionary pressures to be similar across taxa, and yet, research on numerical discrimination has been traditionally focused on mammals, birds, and fishes [@nieder2005counting; @agrillo2014spontaneous]. However, recent studies demonstrate that reptiles are also competent in numbers. For example, red-footed turtles (_Chelonoidis carbonaria_) rapidly learned to associate colour with higher volumes of food [@soldati2017long], Hermann’s tortoises (_Testudo hermanni_) discriminate between large and small food quantities in a spontaneous choice test [@gazzola2018continuous], and freshwater turtles (_Trachemys scripta_ and _Mauremys sinensis_) rapidly learned a quantity discrimination task [@lin2021superior; @lin2024trained]. The Italian wall lizard (_Podarcis siculus_) failed to discriminate between different numbers of food items in a spontaneous choice test [@miletto2017quantitative], but succeeded in a trained quantity discrimination task [@miletto2018quantity]. In contrast, Iberian rock lizards (_Iberolacerta cyreni_) preferred larger quantities of crickets when presented five versus ten items [@recio_prey_2021]. Indeed, the literature shows evidence for basic quantitative abilities in reptiles that may vary with the type of cue and the test employed.

Here, we investigated the effect of prenatal corticosterone (CORT) - the main GC in reptiles - and incubation temperature on numerical discrimination in a lizard. We manipulated CORT concentration (CORT-treated or a sham control) in _L. guichenoti_ eggs and incubated them under one of two temperature regimes (Cold - 23 ± 3ºC or Hot - 28 ± 3ºC) in a 2x2 factorial design. Then, we assessed numerical discrimination abilities of lizards using a spontaneous choice test approach. The numerical discrimination task consisted of five tests that differed in the relative number of food items: 1 VS 4, 1 VS 3, 2 VS 4, 2 VS 3, 3 VS 4 crickets. We based this experiment on observations in nature that the ability to detect a change in stimulus intensity depends on the relative, not absolute, difference between the stimuli ('Weber's law'; @agrillo2014spontaneous). Built upon other studies that explored the cognitive effects of prenatal GCs or thermal early environment in reptiles and other taxa, we hypothesized that lizards incubated at higher temperatures would outperform those incubated at lower temperatures in the numerical discrimination task [@amiel_hotter_2012; @amiel_egg_2014; @clark_colour_2014; @amiel_effects_2017; @vila_pouca_quantity_2019]. We predicted that increased CORT levels would impair numerical competence [@szuran_water_1994; @zhu_prenatal_2004], with more pronounced effects at lower temperatures. Consistent with Weber's law, we predicted that lizards incubated at higher temperatures would successfully discriminate between quantities even at high ratios (e.g. 3 VS 4), whereas even the lowest ratios (e.g. 1 VS 4) could be challenging for those exposed to CORT or incubated at lower temperatures.

@@

All tests were done between 1000-1200 hours when the lizards were most active. Each trial was performed daily for five days between the 17^th^ and 22^nd^ of May 2024. We recorded three main variables: i) _Latency_, defined as the time between the start of the test - when the platform was left in the enclosure - until the first interaction with one of the options; ii) First choice of the lizard (_Choice_), recorded as 1 if the lizard interacted first with the larger amount or 0 if not; and iii) _Interest_ shown for the higher amount of food (detailed below). We considered that the lizard interacted with the food when it touched the Petri dish with its snout or when the front half of the body was on top of the platform. For five minutes after the first interaction with one of the options, we recorded the time spent interacting with that option for a maximum of five minutes; we did this for both amounts of food independently of what was the first choice. We measured the interest shown for the higher amount of food (_Interest_) by subtracting the time interacting with the option with fewer crickets to the time spent interacting with the larger amount of food. If there was no interaction with one of the options during the test, we considered the time spent as zero for that option. When there was no interaction with any of the choices, both _Latency_, _Choice_, and _Interest_ were noted as NA. We also considered Interest as NA in a few cases where the animals managed to lift the dish and eat the prey; Latency and Choice were recorded and employed in the analyses. We excluded from the analyses all the individuals that did not make any choice in three or more tests.

>>>>

## Abstract

**From social behaviour to navigating complex environments, quantitative abilities can be crucial to fitness. However, early life conditions can impact brain development to affect quantitative competence. For example, early thermal conditions or glucocorticoid concentrations (GCs) – stress-related hormones transmitted by the parents – can play a prominent role in shaping cognition through their effects on brain physiology. Furthermore, temperature can lead to increased levels of GCs that can be transmitted to offspring, emphasizing the need to study the combined effects of these two factors. Here, we investigated the effects of elevated prenatal corticosterone (CORT) – the main GC in reptiles – and incubation temperature on numerical discrimination in the common garden skink (_Lampropholis guichenoti_). We assessed numerical abilities through a spontaneous choice test with food as a stimulus. Employing a repeated measures design, we subjected lizards to five numerical tests each differing in the ratios between the two choices (1 VS 4, 1 VS 3, 2 VS 4, 2 VS 3, 3 VS 4). Contrary to our predictions, we found no evidence for treatment effects on lizard behaviour and no use of numerical discrimination during foraging in _L. guichenoti_. Our findings build on previous studies suggesting that reptiles feeding on live prey may have difficulty distinguishing between different prey quantities. Despite our results, further investigation of numerical abilities in live-prey-feeding reptiles and the impact of early conditions on other taxa is warranted, as these areas remain understudied.**

## Significance statement

**Discriminating between quantities can be crucial for individual fitness, yet the impact of early environmental factors on the development of quantitative competence remains largely unexplored. In this study, we investigated, for the first time, the effects of prenatal corticosterone exposure and incubation temperature on numerical discrimination in a reptile. Contrary to our predictions, we did not observe any evidence of numerical discrimination and therefore no impact of early developmental environment. Our results align with previous studies showing that reptiles consuming live prey struggle to differentiate prey numbers in spontaneous choice tests. These results open new avenues for exploring the evolution of numerical abilities in reptiles and identifying factors influencing the development of quantitative skills in this group. Further research should explore alternative methods, such as trained tasks, to better understand the cognitive mechanisms underlying quantitative competence in reptiles.**

@@

The main two approaches experimental approaches in the study of quantity discrimination are spontaneous choice tests and training procedures. In spontaneous choice tests, the subjects are presented two groups of biologically relevant stimuli differing in some quantitative trait (number, size, volumen, etc) and the experimenters record if the subjects select or not the most most appropriate quantity (e.g. the group with more conspecifics or the larger amount of food) (@agrillo2014spontaneous; @cox2016quantity). In training procedures, animals must learn a numerical rule to receive a reward [@agrillo2014spontaneous; @miletto2018quantity]. Both methods are believed to be complementary to each other, and both can provide different results even in the same species [see @miletto2017quantitative and @miletto2018quantity]. One reason is that spontaneous choice tests assume that one of the options presented is more adequate than the other when this may not always be the case. For example, choosing groups of conspecifics with less individuals could be advantageous when food is scarce but disadvantageous when predators are abundant [**REFs first thesis??**]. As such, context and motivation can be important drivers of decision-making that can change the way we measure animals' abilities to discriminate between quantities [@agrillo2014spontaneous]. The other reason is that each method may tap into different cognitive mechanisms [@agrillo2014spontaneous]. In this sense, **Researchers have suggested that numerousness can be processed by two types of systems: one system for small sets (≤ 4) that works by tracking and encoding objects as individual entities; and an analogous system that operates with large numerosities by estimation processes (for a review, see Cantlon et al. (2009); Hyde (2011).**   **HOW CAN EACH BE RELATED TO MOTIVATION OR LEARNING??**

**Differences in test discrete (OFS) versus overall quantity/size (ANS). How to measure both? Creating NUMBER inconsistencies with the SAME SIZE (examples).** OFS, can be challenging, since it is necessary to control for other continuous variables such as area, density, or movement (@agrillo2014spontaneous; **Mehlis et al. 2015**). Generally, authors increase one variable at a time without increases in the other to give conflicting information about the correlation between numerical cues and quantitative continuous traits [**REFs**]. For example, **Look for something where they increase number controlling for size** **HOW??**

No other amphibian or non-avian reptile study has set out to
manipulate these variables to a degree as thoroughly as Stancher et al.42. Future experiments
involving amphibians and non-avian reptiles that manipulate these variables will be crucial for
understanding the relationship between the core systems and the information available for
discrimination

These studies have been traditionally focused on mammals, birds, and fishes despite the influence of numerical cognition on fitness across taxa [@nieder2005counting; @agrillo2014spontaneous]. However, recent studies demonstrate that other vertebrates like amphibians or non-avian reptiles are also competent in numbers [@uller2003salamanders; @stancher2015numerical; @miletto2017quantitative]. For example, red-footed turtles (_Chelonoidis carbonaria_) rapidly learned to associate colour with higher volumes of food [@soldati2017long], Hermann’s tortoises (_Testudo hermanni_) discriminate between large and small food quantities in a spontaneous choice test [@gazzola2018continuous], and freshwater turtles (_Trachemys scripta_ and _Mauremys sinensis_) rapidly learned a quantity discrimination task [@lin2021superior; @lin2024trained]. The Italian wall lizard (_Podarcis siculus_) failed to discriminate between different numbers of food items in a spontaneous choice test [@miletto2017quantitative], but succeeded in a trained quantity discrimination task [@miletto2018quantity]. In contrast, Iberian rock lizards (_Iberolacerta cyreni_) spontaneously preferred larger quantities of crickets when presented five versus ten items [@recio_prey_2021]. **In skinks,**

Lizards were presented with food quantities
differing in number or size and were able to select the larger quantity of food items when they differed in number but not when items
differed in size. We show, for the first time, superior spontaneous discrimination of items differing in number over size in a lizard spe-
cies, which contrasts with previous findings. Our simple method, however, did not include controls for the use of continuous quantities,
and further tests are required to determine the role of such information during quantity discrimination. [@szabo_spontaneous_2021]

We subjected lizards to a Y-maze experiment presenting both discrete (comparisons involving two
vs. four, two vs. six, two vs. eight, four vs. six, four vs. eight, and six vs. eight pieces) and equivalent continuous combinations (single
pieces differing in size) of pumpkin pieces. Our findings revealed that sleepy lizards exhibited spontaneous quantity discrimination, fa-
voring the larger quantity in three discrete comparisons (eight vs. two, two vs. six, four vs. six) and two continuous comparisons (eight
vs. two, eight vs. four). However, low accuracy may be attributable to difficulty in visual acuity exacerbated by difficulty discriminating
quantities at small ratios. [@szabo2024spontaneous]

**These studies demonstrate that reptiles have cognitive abilities related to quantity discrimination that have been overlooked in the past, but that should be explored further.**

Cognitive abilities, however, can be strongly influenced by developmental conditions when the brain is particularly sensitive to environmental inputs [@zhu_prenatal_2004]. For instance, the temperature at which embryos develop can affect neuron density or learning abilities in different species of reptiles [@amiel_hotter_2012; @amiel_egg_2014; @clark_colour_2014; @amiel_effects_2017; @dayananda_incubation_2017; @siviter_incubation_2017; @abayarathna_effects_2020]. These thermal effects on cognition may arise directly or indirectly through impacts on parental condition [@Crino_2023]. Temperatures outside the optimal range can trigger stress responses, leading to the release of glucocorticoids (GCs) [@sapolsky_how_2000], which may then be transmitted to the offspring [@Crino_2023]. Elevated prenatal GCs can, in turn, influence cognition by altering brain function [@szuran_water_1994; @zhu_prenatal_2004; @crino_corticosterone_2014-learn; @farrell_developmental_2015-learn; @bebus_associative_2016; @lui2017chronic]. As such, the interaction between GCs and develpomental temperature must be relevant in shaping reptiles' cognition. Most studies on the effects of GC exposure or elevated temperatures during development focused on learning and memory, but few studies have investigated how early-life conditions can affect quantitative abilities despite the importance of numerical discrimination [but see @vila_pouca_quantity_2019]. Furthermore, the interactive effects of prenatal GCs and temperature on cognitive abilities remain largely unexplored.

Here, we investigated the use of numerical cues in the common garden skink (_Lampropholis guichenoti_) using a spontaneous choice test where lizards were presented with different number of food items while **controlling** for total length. In adition, we examined how prenatal corticosterone (CORT) - the main GC in reptiles - and incubation temperature can affect numerical discrimination and decision-making in foraging contexts. To do so, we manipulated CORT concentration (CORT-treated or a sham control) in _L. guichenoti_ eggs and incubated them under one of two temperature regimes (Cold - 23 ± 3ºC or Hot - 28 ± 3ºC) in a 2x2 factorial design. Then, we assessed numerical discrimination abilities on five tests that differed in the relative number of food items: 1 VS 4, 1 VS 3, 2 VS 4, 2 VS 3, 3 VS 4 crickets.  Previous studies in other skinks showed that these species can discriminater between different quantities of food, although the cues employed to do so remain unclear [@szabo_spontaneous_2021; @szabo2024spontaneous]. We predicted _L. guichenoti_ to be able to discriminate between different number of crickets even when controlling by total length. Furthermore, we predicted lizards' decisions to be influenced by the early conditions they were subjected to. In this regard, we hypothesized that lizards incubated at higher temperatures would outperform those incubated at lower temperatures in the numerical discrimination task [@amiel_hotter_2012; @amiel_egg_2014; @clark_colour_2014; @amiel_effects_2017; @vila_pouca_quantity_2019]. We predicted that increased CORT levels would impair numerical competence [@szuran_water_1994; @zhu_prenatal_2004], with more pronounced effects at lower temperatures. Consistent with Weber's law, we predicted that lizards incubated at higher temperatures would successfully discriminate between quantities even at high ratios (e.g. 3 VS 4), whereas even the lowest ratios (e.g. 1 VS 4) could be challenging for those exposed to CORT or incubated at lower temperatures.

@@

All tests were done between 1000-1200 hours when the lizards were most active. Each trial was performed daily for five days between the 17^th^ and 22^nd^ of May 2024. We recorded three main variables: i) _Latency_, defined as the time between the start of the test - when the platform was left in the enclosure - until the first interaction with one of the options; ii) First choice of the lizard (_Choice_), recorded as 1 if the lizard interacted first with the larger amount or 0 if not; and iii) _Interest_ shown for the higher amount of food (detailed below). We considered that the lizard interacted with the food when it touched the Petri dish with its snout or when the front half of the body was on top of the platform. For five minutes after the first interaction with one of the options, we recorded the time spent interacting with that option for a maximum of five minutes; we did this for both amounts of food independently of what was the first choice. We measured the interest shown for the higher amount of food (_Interest_) by subtracting the time interacting with the option with fewer crickets to the time spent interacting with the larger amount of food. If there was no interaction with one of the options during the test, we considered the time spent as zero for that option. When there was no interaction with any of the choices, both _Latency_, _Choice_, and _Interest_ were noted as NA. We also considered Interest as NA in a few cases where the animals managed to lift the dish and eat the prey; Latency and Choice were recorded and employed in the analyses. We excluded from the analyses all the individuals that did not make any choice in three or more tests. All the analyses were performed by PR, who was blind to the treatment of the lizards.
